---
title: "Analyzing Spotify Extended Streaming History"
author: "Daniel Araujo"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Hello. I was told by a friend that Spotify is able to send you an extended streaming history, and as someone who's excited about data science, I quickly requested my data. My idea with this script is to slowly write code over time, as I have spare time, to analyze the data. Additionally, I plan on peridiocally request the data to see how they change over time.

In case you'd like to try it for yourself, feel free to request your own extended streaming data and copy any chunks of code I wrote. I'm not the most knowledgeable person in R ever, so some code might not be the most efficient. 

## Reading in file

The data is in a .json file, split into many subfiles with the same prefix. So, first I read the files and merged them into a single data frame.

```{r}
library(jsonlite)
library(tidyverse)
library(viridis)
'%&%' = function(a,b) paste (a,b,sep='')

# list of subfiles to read
list_of_json_files <- list.files(path = 'MyData/', pattern = 'endsong')

# read each file and append to a merged data frame
for (f in list_of_json_files){
  tmp <- fromJSON(txt = 'MyData/' %&% f)
  if (exists('full_df')){
    full_df <- rbind(full_df, tmp)
  } else { full_df <- tmp }
}

# only keep columns I want
full_df <- full_df %>% select(ts, ms_played, master_metadata_track_name, 
                        master_metadata_album_artist_name, 
                        master_metadata_album_album_name)
full_df$ms_played <- full_df$ms_played %>% as.numeric()
full_df <- full_df %>% drop_na()
head(full_df)
```


## Modifying timestamp column

The timestamp column (ts) is formatted in the following way: date as YYYY-MM-DD, followed by the time in HH:MM:SS in GMT. However, I want to compare my data across different years/months, so I gotta extract that information from the TS column. 

```{r}
# split timestamp column into multiple
full_df <- full_df %>% separate(col = ts, into = c('year','month','day'), sep = '-')
full_df <- full_df %>% separate(col = day, into = c('day','hour'), sep = 'T')
full_df <- full_df %>% separate(col = hour, into = c('hour','minutes', 'seconds'), sep = ':')
full_df <- full_df %>% select(ms_played, master_metadata_track_name, 
                        master_metadata_album_artist_name,
                        master_metadata_album_album_name,
                        year, month, day, hour)
full_df$year <- full_df$year %>% as.numeric()
full_df$month <- full_df$month %>% as.numeric()
full_df$day <- full_df$day %>% as.numeric()
full_df$hour <- full_df$hour %>% as.numeric()
head(full_df)
```


## Skipped songs 

The original data contains a 'skipped' column, however Spotify doesn't explain well what the values inside that column mean, and most cells were empty. So, I decided to implement my own 'Did I skip this song?' algorithm, which is: if I listened to a song for less than 5 seconds, it's considered as skipped. 

```{r}
# remove songs that I listened to less than 5 seconds
songs_skipped <- full_df %>% filter(ms_played <= 4999)

# see how the dataframe looks like
head(songs_skipped)

# analyze how much I skipped songs
songs_skipped_sum <- songs_skipped %>% group_by(master_metadata_track_name,
  master_metadata_album_artist_name) %>% summarise(times_skipped = n()) %>% 
  unique()
ggplot(songs_skipped_sum, aes(x=times_skipped)) + geom_histogram() + 
  xlab('Number of times I skipped a song') + ylab('Count') + 
  ggtitle('How many times I skip songs')
```

As we can see, most songs I skip few times. Those are probably songs Spotify recommended to me once or twice, and I skipped. However, there are songs I skipped a lot! Those are songs I most likely enjoy, but sometimes I'm not in the mood to listen to them. Let's see the top 10 most skipped songs of all time:

```{r}
# 10 most skipped songs of all time
songs_skipped_sum <- songs_skipped %>% group_by(master_metadata_track_name,
  master_metadata_album_artist_name) %>% summarise(times_skipped = n()) %>% 
  unique()
top10songs_skipped <- songs_skipped_sum %>% arrange(desc(times_skipped)) 
head(top10songs_skipped, n = 10)
```

Bingo! Those are songs I definitely enjoy, but sometimes I just don't wanna listen to them. 

Now, let's see the same data, but this time split across the years. 

```{r}
# analyze how much I skipped songs over the years
songs_skipped_sum <- songs_skipped %>% group_by(master_metadata_track_name,
  master_metadata_album_artist_name, year) %>% summarise(times_skipped = n()) %>% 
  unique()
songs_skipped_sum$year <- songs_skipped_sum$year %>% as.character()
ggplot(songs_skipped_sum, aes(x=times_skipped, fill=year)) + geom_histogram(position = 'dodge') + 
  xlab('Number of times I skipped a song') + ylab('Count') + 
  ggtitle('How many times I skip songs, per year') + scale_fill_viridis_d()
```

If we analyze it by year, the trend is the same. Most songs were skipped once or twice. Also, we can definitely see that I used Spotify during a very brief time in 2015, and didn't ever use it during 2016 and 2017! During this time I used to listen to music using iTunes, so that checks out. 

Let's see the top 10 songs skipped, per year.

```{r}
# 10 most skipped songs per year
songs_skipped_sum <- songs_skipped %>% group_by(master_metadata_track_name,
  master_metadata_album_artist_name, year) %>% summarise(times_skipped = n()) %>% 
  unique()
top10songs_skipped <- songs_skipped_sum %>% arrange(desc(times_skipped)) %>% group_by(year) %>%
  slice(1:10)
top10songs_skipped %>% arrange(desc(times_skipped)) %>% print(n = 1e3)
```

Once again, this confirms how I pretty much didn't use Spotify in 2015, as my top skipped songs were skipped once or twice.  




## RSession

```{r}
sessionInfo()
```

