---
title: "Analyzing Spotify Extended Streaming History"
author: "Daniel Araujo"
date: "Date of data request: August 8th, 2023"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

I learned that, if requested, Spotify sends users their extended streaming history data. Since I'm excited about data science, of course I had to do it. My idea with this script is to slowly write code over time, as I have spare time, to analyze my data. Additionally, I plan on periodically request my streaming data.

In case you'd like to try it for yourself, feel free to request your own extended streaming data and copy any chunks of code I wrote. As a disclaimer, I'm not trying to code the most efficient way in here. It's just supposed to be fun. Don't judge my code! 

---

## 1. Reading input files

The data are in multiple *.json* files with the same prefix. So, first I read the files and merged them into a single data frame, and then removed unnecessary columns. 

```{r}
library(jsonlite)
library(tidyverse)
library(viridis)
library(treemapify)
library(ggrepel)
'%&%' = function(a,b) paste (a,b,sep='')

# list of subfiles to read
list_of_json_files <- list.files(path='MyData/', pattern='Streaming_History_Audio')

# read each file and append to a merged data frame
for (f in list_of_json_files){
  tmp <- fromJSON(txt='MyData/'%&% f)
  if (exists('full_df')){
    full_df <- rbind(full_df, tmp)
    rm(tmp)
  } else { full_df <- tmp }
}

# only keep columns I want
full_df <- full_df %>% select(master_metadata_album_artist_name, master_metadata_track_name, ts, ms_played) %>% drop_na()

# make sure the timestamp column is in the correct format
full_df$ts <- as_datetime(full_df$ts)

# renaming first two columns (names are too big!)
colnames(full_df)[1:2] <- c('artist_name','track_name')
```

---

## 2. Skipped songs 

The original data contains a 'skipped' column with Boolean values, however most cells were empty. Thus, I decided to implement my own 'Did I skip this song?' algorithm, which basically is: if I listened to a song for less than 10 seconds, it's considered as skipped. 

```{r}
# keeps songs that I listened to for less than 10 seconds
songs_skipped <- full_df %>% filter(ms_played<=9999)

# analyze how much I've skipped songs
songs_skipped_summary <- songs_skipped %>% group_by(track_name, artist_name) %>% 
  summarise(times_skipped=n()) %>% unique()

# plot a histogram
ggplot(songs_skipped_summary, aes(x=times_skipped)) + geom_histogram() + 
  xlab('Number of times I have skipped a song') + ylab('Count') + 
  ggtitle('How many times I have skipped songs') + theme_minimal()
```

As we can see, most songs I have skipped few times. Those are probably songs Spotify recommended to me once or twice, and I skipped them. However, there are songs I have skipped a lot! Let's see the top 10 most skipped songs of all time:

```{r}
# 10 most skipped songs of all time
songs_skipped_summary %>% arrange(desc(times_skipped)) %>% head(n=10)
```

Bingo! Those are songs I definitely enjoy, but sometimes I just don't wanna listen to them. 

---

## 3. Most streamed songs 

OK, so now that I was able to briefly analyze the data corresponding to my skipped songs, let's take a look at the ones I actually listened to. First, let's see my top 100 most streamed songs of all my time as a Spotify user.  

```{r}
# keeps songs that I listened to for at least 10 seconds
songs_listened <- full_df %>% filter(ms_played>=10000)

# analyze how much I've streamed songs
songs_listened_summary <- songs_listened %>% group_by(track_name, artist_name) %>% 
  summarise(times_streamed=n()) %>% unique()

# 100 most streamed songs of all time
songs_listened_summary %>% arrange(desc(times_streamed)) %>% print(n=100)
```

I feel like any comment about my top 100 list would be TMI. It is what it is.

Now, let's see the top 10 songs per year. 

```{r}
# sum amount of times I've streamed each song per year
songs_listened_summary <- songs_listened %>% group_by(track_name, artist_name, year(ts)) %>% 
  summarise(times_streamed=n()) %>% unique()

# rename "year(ts)" column
colnames(songs_listened_summary)[3] <- c('year')

# get top 10 per year
top10songs_listened <- songs_listened_summary %>% group_by(year) %>% arrange(desc(times_streamed)) %>% 
  slice(1:10)

# print full data frame
top10songs_listened %>% arrange(desc(times_streamed)) %>% print(n=nrow(top10songs_listened))
```

It is interesting to see my top 10 songs of each year, because I definitely see how it changes over the years. It is also possible to notice that I basically did not use Spotify between 2015-2017, so I will make sure to remove those years from my data frame. 

```{r}
# only keep streams that occurred in 2018 or after
songs_listened <- songs_listened %>% filter(year(ts)>2017)
```

---

## 4. Favorite artists

Now, what are the artists I listen to the most? I have my own guesses, but let's see what the data tell us. 

```{r}
# get the total amount of times I've listened to each artist
artists_frequency <- songs_listened %>% group_by(artist_name) %>% 
  summarise(times_listened=n())

# print the top 100
artists_frequency %>% arrange(desc(times_listened)) %>% print(n=100)
```

Am I surprised Lana Del Rey is in the first place? No. Am I shocked by the difference between her and the second place? A bit. But overall, the top 100 fairly represents my musical taste (duh!). 

Let's do top 10 per year now. 

```{r}
# get the total amount of times I've listened to each artist per year
artists_frequency <- songs_listened %>% group_by(artist_name, year(ts)) %>% 
  summarise(times_listened=n())

# rename "year(ts)" column
colnames(artists_frequency)[2] <- c('year')

# get top 10 per year
top10artists_listened <- artists_frequency %>% group_by(year) %>% arrange(desc(times_listened)) %>% 
  slice(1:10)

# print full data frame
top10artists_listened %>% arrange(desc(times_listened)) %>% print(n=nrow(top10artists_listened))
```

I feel like the breakdown per year does not have as much information as the previous one, but it is possible to see how in some years I was more into certain music genres than others. 

Just out of curiosity, I would like to know how often the same artists are found within my top 10 across all years. 

```{r}
# count how many times an artist appears in a top 10
top10artists_frequency <- top10artists_listened %>% group_by(artist_name) %>% summarise(times_in_top10=n())

# make a treemap
ggplot(top10artists_frequency, aes(area=times_in_top10, fill=times_in_top10, label=artist_name, subgroup=times_in_top10)) + labs(fill='Artist') + geom_treemap() + geom_treemap_text() + geom_treemap_subgroup_border(color='black') + geom_treemap_subgroup_text(place='centre', grow=T, alpha=0.6) + theme(legend.position='none') + scale_fill_viridis()
```

In the treemap above, gray numbers represent how many times the artists in each subgroup (same color) are found in a top 10. Lana Del Rey is the only artist that has appeared in a top 10 every year. No surprises there. 

---

## 5. Artist revenue

Contrary to what most people believe, Spotify does not pay artist royalties according to a per-play or per-stream rate. It is more complicated than that. However... what if they did? According to a totally not in-depth and scientific research I did, most artists are paid roughly 0.004 USD per stream. So using a little bit of math, let's see how much money artists have made from my streams alone.

```{r}
# get the total amount of times I've listened to each artist and multiply by the "payment rate"
artists_revenue <- songs_listened %>% group_by(artist_name) %>% 
  summarise(revenue=n()*0.004)

# print the top 100
artists_revenue %>% arrange(desc(revenue)) %>% print(n=100)
```

This result honestly made me laugh. The revenue values are very low. However, I am just one person, right? Given the huge amount of Spotify users, artists make much more money than that. Additionally, they have other sources of income as well.  

---

## 6. Time spent listening to music

For this section, I would like to know how many days per year I spent listening to music. 

```{r}
# make a 'days_played' column based on the 'ms_played' column
# 1 millisecond = 1.15741e-8 days
songs_listened <- songs_listened %>% mutate(days_played=ms_played*1.15741e-8)

# for every year, sum the days_played column
time_spent_w_songs <- songs_listened %>% group_by(year(ts)) %>% summarise(days_listened=sum(days_played))

# rename "year(ts)" column
colnames(time_spent_w_songs)[1] <- c('year')

# let's plot it!
ggplot(time_spent_w_songs, aes(x=year, y=days_listened)) + geom_col() + 
  xlab('Year') + ylab('Number of days') + ggtitle('Days spent listening to music in each year') + theme_minimal()
```

So many days listening to music in 2020! I blame COVID-19. 

---

## 7. Time spent streaming certain artists

Now that we know how many days per year I have spent streaming songs, let's do something similar but now with the artists in my top 10 overall. 

```{r}
# get the names of the artists in my top 10
top10artists <- songs_listened %>% group_by(artist_name) %>% 
  summarise(times_listened=n()) %>% arrange(desc(times_listened)) %>% 
  slice(1:10) %>% pull(artist_name)

# filter main data frame so it only contains artists in my top 10
top10artists_streamingtime <- songs_listened %>% filter(artist_name %in% top10artists)

# compute total streaming time
top10artists_streamingtime <- top10artists_streamingtime %>% group_by(artist_name) %>% 
  summarise(streaming_time=sum(days_played))

# plot
ggplot(top10artists_streamingtime, aes(x=reorder(artist_name, -streaming_time), y=streaming_time)) + 
  geom_col() + coord_flip() + xlab('Artist') + ylab('Number of days') + 
  ggtitle('Days spent listening to music by artists in my top 10') + theme_minimal()
```

I must admit, that is a lot of days listening to Lana Del Rey. But the results look so good visualized this way!  

---

## 8. Favorite artists per week

I wanna try and plot what were my top artists per week, across all my years as a Spotify user. What I hope to see is how my most listened to artists per week change according to some life events, such as concerts, plays, new album releases, etc. 

```{r, fig.width=10, fig.height=70}
# for every year and week, sum the time spent listening to every artist
topartists_week <- songs_listened %>% group_by(year(ts), week(ts), artist_name) %>% summarise(time=sum(ms_played))

# rename lubridate-made columns
colnames(topartists_week)[1:2] <- c('year', 'week')

# for every year and week, select the top 3 artists and add a new column with their respective position in the top 3
topartists_week <- topartists_week %>% group_by(year, week) %>% slice_max(time, n=3) %>% mutate(position=c(1:3))

# plot it!
ggplot(topartists_week, aes(x=week, y=position, color=artist_name, label=artist_name)) + geom_point(show.legend=F) + geom_line(show.legend=F) + geom_label_repel(show.legend=F) + facet_wrap(~year, ncol=1) + scale_y_continuous(breaks=c(1,2,3)) + scale_x_reverse(breaks=c(1,10,20,30,40,50)) + coord_flip() + xlab('Week') + ylab('Rank') + ggtitle('Top 3 artists per week in every year') + theme_minimal()
```

Ok... this isn't my most beautiful work. The plot is too busy, but I don't think I could improve it further. Anyway, regardless of how busy the figure is, it's still possible to draw some information from it. For instance, we can see the influence in my top artists of the week in weeks in which I saw some musicals in the theater (such as SIX on May 10th, 2022), an artists released a new album (for example, Dua Lipa's Future Nostalgia on March 27th, 2020 and Beyoncé's Renaissance on July 29th, 2022), or I first discovered an artist for the first time and streamed them a bunch (such as Weathers on September 6th-12th).

Overall, the plot tells an interesting story. But to be fair, this plot is probably only that interesting to me. 

---

## RSession

```{r}
sessionInfo()
```

